---
title: "Beta Null Deviations"
author: "John Quensen"
date: "February 4, 2017"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Beta Null Deviations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Make a small community matrix of counts to use as an example. Species are in columns and samples are in rows.  

```{r}
suppressMessages(library(vegan))
data("dune")
comm <- dune
comm <- comm[1:5, ]
comm <- comm[ , colSums(comm)>0]
comm <- comm[ , 1:8]
comm
```

Individuals of each species are shuffled among samples by method "`c0_ind`." One way of generating a series of randomized communities by this method is to first define the null model with `nullmodel` and then replicate it a number of times with `simulate` as in the commands:  

```{r}
x1 <- nullmodel(comm, "c0_ind")
x2 <- simulate(x1, nsim = 999)
x2
```

The object `x2` contains 999 randomized matrices. In order to compare with the original `comm`, the first randomized matrix can be displayed as:  

```{r}
x2[ , , 1]
```

The column sums of `comm` and each of the randomized matrices are the same. If we consider each row as comprising a patch in a metacommunitiy, then the randomization procedure maintians the number of individuals in each species in the metacommunity.  

```{r}
colSums(comm)
colSums(x2[ , , 1])
```

Consequently, the total number of individuals and the number of non-empty species ($\gamma$-diversity) are held constant.  

```{r}
sum(comm)
sum(x2[ , , 1])

sum(colSums(comm)>0)
sum(colSums(x2[ , , 1])>0)
```

To check that there are no empty species in any of the replications:  

```{r}
all(apply(x2, 3, colSums)>0)
```

But the number of individuals per sample ($\alpha$-diversity) is not held constant between replicates.  

```{r}
rowSums(comm)
rowSums(x2[ , , 1])
```

Null models are sometimes used to generate expected $\beta$-diversites if $\alpha$- and $\gamma$-diversities are held constant on replication, but that is not the case here.  

```{r}
bray.obs <- mean(vegdist(comm, "bray"))
bray.nulls <- apply(x2, 3, vegdist, "bray")
bray.exptd <- mean(bray.nulls)
bray.expt.sd <- sd(bray.nulls)
ses <- (bray.obs - bray.exptd)/bray.expt.sd
bray.obs
bray.exptd
bray.expt.sd
ses
```

Plot the distribution of expected mean bray distances. Place the observed mean bray in that distribution.  

```{r}
hist(bray.nulls, main = "Bray-nulls Distribution")
abline(v = bray.obs, col = "red")
# Find the probablility that the observed mean bray distance
# is not significantly different from the expected mean bray distance.
rank.1 <- rank(c(bray.obs, bray.nulls))[1]
p.val <- rank.1/(length(bray.nulls)+1)
p.val
```

Compare mean observed Bray distance to mean null Bray distance using `oecosim`.  

```{r}
mean.bray <- function(x) mean(vegdist(x, "bray"))
rslt <- oecosimu(comm, mean.bray, method = "c0_ind", nsimul = 999)
rslt
```

